Parsing Postfix log files

### 2007-11-11 doc/logparser.tex
More acronyms. Wrapped all Postfix programs except sendmail in \daemon{}. Added glossary entries for all Postfix programs, some need to be fleshed out. Some corrections. More references.

@@

    \item [regex] The regular expression~\cite{wikipedia-regex, perlre} to
        match the log line against.  The regex will first have several
        keywords expanded: this simplifies reading and writing rules;
        avoids needless repetition of complex regex components; allows the
        components to be corrected and/or improved in one location; and
        makes each regex largely self-documenting.

@@

        For efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file --- otherwise each
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown.  Rule efficiency concerns are
        discussed in section~\refwithpage{rule efficiency}.

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is: \newline
        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline i.e.\
        semi-colon or comma separated assignment statements, with the
        variable name on the left and the matching field from the regex on
        the right hand side.  The list of acceptable variable names is:

@@

    \item [queueid] Specifies the matching field from the regex which gives
        the queueid, or zero if the log line doesn't contain a queueid.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate a \DFA{} from it, and examine
every \DFA{} to determine if any overlap.  The author has not attempted to
do this: such a project by itself would probably qualify for a PhD, and may
involve solving the Halting Problem and circumventing the Church-Turing
Thesis.  XXX REFERENCES FOR HALTING PROBLEM AND CHURCH-TURING\@.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line is unspecified, and
the first matching rule wins.  Overlapping rules are frequently a
requirement, allowing a more specific rule to match some lines and a more
general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

@@

    \item Sort the regexs and visually inspect the list, e.g.\ with SQL
        similar to: \textbf{select regex from rules order by regex;}

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line the regex will be tried against the
        line; if the match is successful the action will be executed, if
        not the next rule will be tried.  If the program-specific rules
        don't match the log line, the generic rules will be tried in the
        same way.

    \item [action:] will be executed if the regex matches successfully (see
        section~\refwithpage{actions-in-detail} for full details).

@@

    \item [queueid:] The index of the field in the regex which captured the
        queueid, or zero if the line does not contain a queueid.  This
        allows the correct mail can be found by queueid and actions
        performed on it.

@@

Perl compiles the original regex into an internal representation,
optimising the regex to improve the speed of matching, but this compilation
and optimisation takes CPU time.  Perl automatically caches static regexs,
but dynamic regexs need to be explicitly compiled and cached.
Graph~\refwithpage{normal regex vs discard regex} shows execution times
with and without caching the regex.  Caching the compiled regexs is
obviously far more efficient; graph~\refwithpage{normal regex vs discarded
regex factor} shows the percentage execution time increase when not caching
each regex.

Caching the compiled regexs is quite simple, and is the single most
effective optimisation implemented in the parser.

@@

    \item The final part of this section discusses rule efficiency,
        covering the effects of caching compiled regexs and optimal
        ordering of rules, with reference to the graphs in
        appendix~\refwithpage{graphs}.

@@

The mail lacks a queueid until it has been accepted, so log lines must
first be correlated by the smtpd pid, then transition to being correlated
by the queueid.  This is relatively minor, but does require:

@@

    \item Most importantly: every section of code must know whether it
        needs to lookup the data structures by pid or queueid.

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by pid for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see section~\refwithpage{timeouts-during-data-phase} for details.
However in general there isn't any ambiguity about which data structure
should be used for a given log line.}  The original connection will be
discarded unsaved when the client disconnects if it doesn't have any data
worth saving, i.e.\ no rejections.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are addressed to and accepted for a local address but need to be
delivered to one or more remote addresses due to aliases.  When this occurs
a child mail will be injected into the Postfix queue, but without the
explicit logging smtpd or postdrop injected mails have.  Thus the source is
not immediately discernible from the log line in which the mail first
appears; from a strictly chronological reading of the logs it
\textit{usually\/} appears as if the child mail has appeared from thin air.
Subsequently the parent mail will log the creation of the child mail:

@@

Unfortunately while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition the first log line belonging to the child mail (the log line cited
above properly belongs to the parent mail) is logged by qmgr,\footnote{Qmgr
is the Postfix daemon which manages the mail queue, determining which mails
will have delivery attempted next.} so the order also depends on how busy
qmgr is.\footnote{Postfix is quite paranoid about mail delivery, an
excellent characteristic for an \MTA{} to possess, so it won't log that the
child has been created until it is absolutely certain that the mail has
been written to disk.}

Because of this the parser cannot complain when it encounters a log line
from qmgr for a previously unseen mail; it must flag the mail as coming
from an unknown origin, and subsequently clear the flag if and when the
origin of the mail becomes clear.  Obviously the parser could omit checking
of where mails originate from, but the author believes that it is better to
require an explicit source, as bugs in the parser are more likely to be
exposed.

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.  The parser thus requires both a method of mapping log
lines to the correct mail and a method of specifying the action the log
line represents.  The former is achieved by using the smtpd process id to
identify the correct mail during the initial phase, then switching to the
queueid once the mail has been accepted.  The latter uses the action field
of the rule which matched the log line, executing the code in the function
named by the action.

@@

Everything starts off with a mail entering the system, whether by local
submission via postdrop/sendmail,\footnote{Postfix provides a command that
is compatible with the Sendmail (\url{http://www.sendmail.org/ XXX CHECK})
mail submission program which all Unix commands which send mail depend on.}
by \SMTP{}, by re-injection due to forwarding, or internally generated by
Postfix.  Local submission is the simplest case: a queueid is assigned
immediately and the sender address is logged (action: pickup;
flowchart:~2).

@@

    \item The last alternative is that the mail is a parent mail (action:
        commit; flowchart:~24).  Regardless of the state of its children
        the data is cleaned up and entered in the database (flowchart:~25).
        The parent may have children which are waiting to be entered in the
        database (flowchart:~26); each of those children's data is cleaned
        up and entered in the database, then deleted from the state tables.
        The parent may also have outstanding children which are not yet
        delivered, in which case (flowchart:~27) the parent must wait for
        those children to be finished with.  As soon as the last child is
        deleted from the state tables the parent will also be finished with
        (flowchart:~28), and deleted from the state tables.

@@

    \item [\$line] The log line, with the standard syslog(3) fields parsed,
        removed and made available.

    \item [\$rule] The matching rule.

    \item [@matches] The fields in the line captured by \$rule's regex.

@@

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by smtpd pid) and save both the client
        hostname and IP address.

@@

    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
        the log line, and save the data extracted by the regex to it.

@@

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both qmgr and cleanup as it needs to deal with out of order log
        lines; see section~\refwithpage{discarding cleanup lines} for
        details.

    \item [PICKUP] Pickup is the service which deals with mail submitted
        locally via /usr/sbin/sendmail.

        Out of order log entries may have caused the state table entry to
        already exist (see section~\refwithpage{pickup logging after
        cleanup}); otherwise it is created.  The data from the log line is
        then saved to the state table entry.

@@

    \item [MAIL\_TOO\_LARGE] When the client tries to send a message larger
        than the local server accepts the mail will be discarded.  See
        TIMEOUT for further discussion; the two are handled in exactly the
        same way.

@@

    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
        it kills all smtpds,\footnote{Possibly other programs are killed
        also, but the parser is only affected by and interested in smtpds
        exiting.} requiring any active connections to be committed and
        deleted from the state tables.

    \item [SMTPD\_KILLED] Sometimes Postfix needs to forcefully kill an
        smtpd; the active connection for that smtpd must be committed and
        deleted from the state tables.

    \item [SMTPD\_DIED] Sometimes an smtpd dies or exits unsuccessfully;
        the active connection for that smtpd must be committed and deleted
        from the state tables.

    \item [SMTPD\_WATCHDOG] Smtpds have a watchdog timer to deal with
        unusual situations --- after five hours the timer will expire and
        the smtpd will exit.  This occurs very infrequently, as there are
        many other timeouts which should occur in the intervening hours:
        DNS timeouts, timeouts reading data, etc.  The active connection
        for that smtpd must be committed and deleted from the state tables.

    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
        creation of bounce messages.  This action creates a new mail if
        necessary, marking it as a bounce notification and removing the
        unknown origin flag if it already exists.

@@

    \item Neither smtpd nor pickup have logged any messages associated with
        the mail, indicating it was generated internally by Postfix, not
        accepted via \SMTP{} or submitted locally by sendmail.

@@

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing.  In the later case the message-id will be
        unchanged from the original bounce notification, and so even if it
        happens to be in the correct format (e.g.\ if it was generated by
        Postfix on another server) the queueid in the message-id will not
        match the queueid of the mail.

@@

Once a mail has been identified as a bounce notification the unknown origin
flag is cleared and the mail can be entered in the database.

@@

The most likely cause of mis-identification is if a mail generated
internally by Postfix is identified as a bounce notification when it is a
different type of message; arguably this is a benefit rather than a
drawback, as future mails generated internally by Postfix will be handled
correctly.

@@

    \item Is the mail missing its cleanup log message?  Every mail which
        passes through Postfix will have a cleanup line; lack of a cleanup
        line is a sure sign the mail didn't make it too far.

    \item Were there exactly two smtpd log lines for the mail?  There
        should be a connection line and a mail accepted line.

@@

    \item Is the mail missing its cleanup log message?  Every mail which
        passes through Postfix will have a cleanup line; lack of a cleanup
        line is a sure sign the mail didn't make it too far.

    \item Were there three or more smtpd log lines for the mail?  There
        should be a connection line and a mail accepted line, followed by
        one or more rejection lines.

    \item Is the last smtpd log line a rejection line?

@@

        A timeout may thus apply either to an accepted mail or a rejected
        mail.  To distinguish between the two cases the algorithm compares
        the timestamp of the last accepted mail against the timestamp of
        the last line logged by smtpd for that connection.  If the smtpd
        timestamp is later there was a rejection between the accepted mail
        and the timeout, therefore the timeout applies to a rejected mail;
        the timeout has already been recorded so the timeout action
        finishes.  If the mail acceptance timestamp is greater then the
        timeout applies to the just-accepted mail, which will be discarded.

@@

This complication is further complicated by out of order cleanup lines: see
section~\refwithpage{discarding cleanup lines} for details.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances which trigger it.  Sometimes the
cleanup line is logged after the timeout line; parsing this line causes the
creation of a new state table entry for the queueid in the log line.  This
is incorrect because the line actually belongs to the mail which has just
been discarded, and the next log line for that queueid will be seen when
the queueid is reused, causing a queueid clash and the appropriate warning.

In the case where the cleanup line is still pending the algorithm updates a
global cache of queueids, adding the queueid and the timestamp from the
timeout line.  When the next cleanup line is parsed for that queueid the
cache will be checked, and the line will be deemed part of the discarded
mail and discarded if it meets the following requirements:

@@

    \item The timestamp of the cleanup line must be within ten minutes of
        the mail acceptance timestamp.  Timeouts happen after five minutes,
        but some data may have been transferred slowly, and empirical
        evidence shows that ten minutes is not unreasonable; hopefully it
        is a good compromise between false positives and false negatives.

@@

The next cleanup line must meet the criteria above for it to be discarded
because not every connection where a timeout occurs will have a cleanup
line logged for it; if the algorithm blindly discarded the next cleanup
line it would in some cases be mistaken.  Whether or not the next cleanup
line is discarded the queueid will be removed from the cache of timeout
queueids when the next pickup line containing that queueid is parsed.

During the TIMEOUT action if the mail lacks a cleanup line, the queueid and
other data about the mail, particularly the timestamp of the last log line,
are added to a global cache of timeout queueids.  This cache is
subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action, and
if the queueid exists and the cleanup line passes the checks listed above
it will be discarded.

@@

Occasionally the pickup line logged when mail is submitted locally via
sendmail appears later in the logfile than the cleanup line for that mail.
This seems to occur during periods of particularly heavy load, so is most
likely due to process scheduling vagaries.  Normally if the queueid given
in the pickup line exists a warning is generated by the pickup action, but
if the following conditions are met it is assumed that the lines are out of
order:

@@

    \item The only program which logged anything for the mail is cleanup.

    \item There is less than a five second difference in the timestamps of
        the cleanup and pickup lines.

@@

Occasionally an smtpd will just stop logging, without an immediately
obvious reason.  After poring over logs for some time there are several
reasons for this infrequent occurrence:

@@

    \item Postfix is stopped or its configuration is reloaded.  When this
        happens all smtpds exit, and all entries in the connections state
        table must be committed and deleted from the state table.  The
        queueid state table is untouched.

    \item Postfix sometimes forcibly kills an smtpd, so the active
        connection must be committed and deleted from the connections state
        table.

    \item Occasionally an smtpd will die or exit uncleanly, so the active
        connection must be committed and deleted from the connections state
        table.

@@

The above appear to cover all situations where an smtpd suddenly stops
logging.  In addition to removing an active connection the last accepted
mail may need to be discarded, as detailed in
section~\refwithpage{timeouts-during-data-phase}.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in section~\refwithpage{additional complications}
were detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
pid is not found,\footnote{There will often be warnings about missing
queueids or pids in the first few hundred or thousand log lines because the
earlier log lines for those connections or mails are in the previous log
file; loading the saved state from the previous log file will solve this
problem.} or when there are problems tracking a child mail (see
section~\refwithpage{tracking re-injected mail}).  There should be few or
no warnings, and the state table should only contain entries for mails
which had yet to be delivered when the logfiles ended, or started before
the logfiles began.

@@

        Pflogsumm produces a report designed for troubleshooting, and does
        not support saving any data; parsing is fixed and concentrates on
        extracting only the data required to produce the report.  Both the
        parsing and reporting are difficult to extend, as it is a
        specialised tool.  It does not attempt to correlate log lines by
        queueid or pid, nor does it need to deal with the complications
        encountered during this project.  Pflogsumm produces a useful
        report, and successfully dealt with all the Postfix logs tested
        with.

@@

        The interface is optimised for interactive rather than automated
        queries and it does not appear to be possible to write independent
        tools to utilise the Splunk database.  Splunk is a generic tool, so
        it lacks any Postfix specific support over and above extracting the
        \texttt{key=<value>} fields; most importantly it makes no attempt
        to correlate log lines by queueid or pid, nor to handle any of the
        myriad complications discussed in this document.  Correlation of
        log lines is left up to the user, and Splunk certainly makes the
        job easier than using grep, awk and similar Unix tools unless the
        user has a very high proficiency with those tools.

@@

        Unlike other parsers reviewed, Lire does attempt to correlate log
        lines by queueid, but not by pid, so e.g.\ the complete list of
        recipients for a mail should be available.

>>>>

% This is ridiculous, but I can't put @ in glossary entries, so . . .
\newcommand{\at}[0]{%
    @%
}

@@

    \item [regex] The \regex{} to match the log line against.  The \regex{}
        will first have several keywords expanded: this simplifies reading
        and writing rules; avoids needless repetition of complex \regex{}
        components; allows the components to be corrected and/or improved
        in one location; and makes each \regex{} largely self-documenting.

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Rule efficiency concerns are
        discussed in section~\refwithpage{rule efficiency}.

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is: \newline
        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline i.e.\
        semi-colon or comma separated assignment statements, with the
        variable name on the left and the matching field from the \regex{}
        on the right hand side.  The list of acceptable variable names is:

@@

    \item [queueid] Specifies the matching field from the \regex{} which
        gives the queueid, or zero if the log line doesn't contain a
        queueid.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's \regex{}, generate a \DFA{} from it, and
examine every \DFA{} to determine if any overlap.  The author has not
attempted to do this: such a project by itself would probably qualify for a
PhD, and may involve solving the Halting Problem and circumventing the
Church-Turing Thesis.  XXX REFERENCES FOR HALTING PROBLEM AND
CHURCH-TURING\@.} that responsibility is left to the author of the rules.
Unintentionally overlapping rules lead to inconsistent parsing and data
extraction because the order in which rules are tried against each line is
unspecified, and the first matching rule wins.  Overlapping rules are
frequently a requirement, allowing a more specific rule to match some lines
and a more general rule to match the majority, e.g.\ separating \SMTP{}
delivery to a specific sites from \SMTP{} delivery to the rest of the
world.  The algorithm provides a facility for ordering overlapping rules:
the priority field in each rule (defaults to zero).  Rules are sorted by
priority, highest first, and then rules with the same priority are sorted
by the number of successful matches when parsing the previous log file.
Negative priorities may be useful for catchall rules.

@@

    \item Sort by regex and visually inspect the list, e.g.\ with SQL
        similar to: \textbf{select regex from rules order by regex;}

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line the \regex{} will be tried against
        the line; if the match is successful the action will be executed,
        if not the next rule will be tried.  If the program-specific rules
        don't match the log line, the generic rules will be tried in the
        same way.

    \item [action:] will be executed if the \regex{} matches successfully
        (see section~\refwithpage{actions-in-detail} for full details).

@@

    \item [queueid:] The index of the field in the \regex{} which captured
        the queueid, or zero if the line does not contain a queueid.  This
        allows the correct mail can be found by queueid and actions
        performed on it.

@@

Perl compiles the original \regex{} into an internal representation,
optimising the \regex{} to improve the speed of matching, but this
compilation and optimisation takes CPU time.  Perl automatically caches
static \regexs{}, but dynamic \regexs{} need to be explicitly compiled and
cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
times with and without caching the regex.  Caching the compiled \regexs{}
is obviously far more efficient; graph~\refwithpage{normal regex vs
discarded regex factor} shows the percentage execution time increase when
not caching each regex.

Caching the compiled \regexs{} is quite simple, and is the single most
effective optimisation implemented in the parser.

@@

    \item The final part of this section discusses rule efficiency,
        covering the effects of caching compiled \regexs{} and optimal
        ordering of rules, with reference to the graphs in
        appendix~\refwithpage{graphs}.

@@

The mail lacks a queueid until it has been accepted, so log lines must
first be correlated by the \daemon{smtpd} \pid{}, then transition to being
correlated by the queueid.  This is relatively minor, but does require:

@@

    \item Most importantly: every section of code must know whether it
        needs to lookup the data structures by \pid{} or queueid.

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by \pid{} for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see section~\refwithpage{timeouts-during-data-phase} for details.
However in general there isn't any ambiguity about which data structure
should be used for a given log line.}  The original connection will be
discarded unsaved when the client disconnects if it doesn't have any data
worth saving, i.e.\ no rejections.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are addressed to and accepted for a local address but need to be
delivered to one or more remote addresses due to aliases.  When this occurs
a child mail will be injected into the Postfix queue, but without the
explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
Thus the source is not immediately discernible from the log line in which
the mail first appears; from a strictly chronological reading of the logs
it \textit{usually\/} appears as if the child mail has appeared from thin
air.  Subsequently the parent mail will log the creation of the child mail:

@@

Unfortunately while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition the first log line belonging to the child mail (the log line cited
above properly belongs to the parent mail) is logged by \daemon{qmgr}, so
the order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is
quite paranoid about mail delivery, an excellent characteristic for an
\MTA{} to possess, so it won't log that the child has been created until it
is absolutely certain that the mail has been written to disk.}

Because of this the parser cannot complain when it encounters a log line
from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
coming from an unknown origin, and subsequently clear the flag if and when
the origin of the mail becomes clear.  Obviously the parser could omit
checking of where mails originate from, but the author believes that it is
better to require an explicit source, as bugs in the parser are more likely
to be exposed.

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.  The parser thus requires both a method of mapping log
lines to the correct mail and a method of specifying the action the log
line represents.  The former is achieved by using the \daemon{smtpd}
process id to identify the correct mail during the initial phase, then
switching to the queueid once the mail has been accepted.  The latter uses
the action field of the rule which matched the log line, executing the code
in the function named by the action.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
due to forwarding, or internally generated by Postfix.  Local submission is
the simplest case: a queueid is assigned immediately and the sender address
is logged (action: pickup; flowchart:~2).

@@

    \item The last alternative is that the mail is a parent mail (action:
        commit; flowchart:~24).  Regardless of the state of its children
        its data is cleaned up and entered in the database (flowchart:~25).
        The parent may have children which are waiting to be entered in the
        database (flowchart:~26); each of those children's data is cleaned
        up and entered in the database, then deleted from the state tables.
        The parent may also have outstanding children which are not yet
        delivered, in which case (flowchart:~27) the parent must wait for
        those children to be finished with.  As soon as the last child is
        deleted from the state tables the parent will also be finished with
        (flowchart:~28), and deleted from the state tables.

@@

    \item [line] The log line, separated into fields:

            \item [timestamp] The time the line was logged at.

            \item [host] The hostname of the server which logged the line.

            \item [program] The name of the program which logged the line.

            \item [pid] The \pid{} of the program which logged the line.

            \item [text] The remainder of the line.

    \item [rule] The matching rule.

    \item [matches] The fields in the line captured by the rule's regex.

@@

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
        client hostname and IP address.

@@

    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
        the log line, and save the data extracted by the \regex{} to it.

@@

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
        out of order log lines; see section~\refwithpage{discarding cleanup
        lines} for details.

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  Out of order log
        entries may have caused the state table entry to already exist (see
        section~\refwithpage{pickup logging after cleanup}); otherwise it
        is created.  The data from the log line is then saved to the state
        table entry.

@@

    \item [MAIL\_TOO\_LARGE] When the client tries to send a message larger
        than the local server accepts, the mail will be discarded, and the
        client informed.  See TIMEOUT for further discussion; the two are
        handled in exactly the same way.

@@

    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
        it kills all \daemon{smtpd} processes,\footnote{Possibly other
        programs are killed also, but the parser is only affected by and
        interested in \daemon{smtpd} processes exiting.} requiring any
        active connections to be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [SMTPD\_KILLED] Sometimes an event occurs which requires Postfix
        to forcefully kill a \daemon{smtpd}; that \daemon{smtpd} processes'
        active connection must be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
        unsuccessfully; the active connection for that \daemon{smtpd} must
        be cleaned up, entered in the database, and deleted from the state
        tables.

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations --- after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts which should occur
        in the intervening hours: \DNS{} timeouts, timeouts reading data
        from the client, etc.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
        creation of bounce messages.  This action creates a new mail if
        necessary, marking it as a bounce notification and removing the
        unknown origin flag if it exists.

@@

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating it was generated
        internally by Postfix, not accepted via \SMTP{} or submitted
        locally by \daemon{postdrop} or sendmail.

@@

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing.  In the latter case the message-id will be
        unchanged from the original bounce notification, and so even if it
        happens to be in the correct format (e.g.\ if it was generated by
        Postfix on another server) the queueid in the message-id will not
        match the queueid of the mail.

@@

Once a mail has been identified as a bounce notification the unknown origin
flag is cleared and the mail can be cleaned up and entered in the database.

@@

The most likely cause of mis-identification is if a mail generated
internally by Postfix is identified as a bounce notification when it is a
different type of message; arguably this is a benefit rather than a
drawback, as other mails generated internally by Postfix will be handled
correctly.

@@

    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
        which passes through Postfix will have a \daemon{cleanup} line;
        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
        it too far.

    \item Were there exactly two \daemon{smtpd} log lines for the mail?
        There should be a connection line and a mail accepted line.

@@

    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
        which passes through Postfix will have a \daemon{cleanup} line;
        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
        it too far.

    \item Were there three or more \daemon{smtpd} log lines for the mail?
        There should be a connection line and a mail accepted line,
        followed by one or more rejection lines.

    \item Is the last \daemon{smtpd} log line a rejection line?

@@

        A timeout may thus apply either to an accepted mail or a rejected
        mail.  To distinguish between the two cases the algorithm compares
        the timestamp of the last accepted mail against the timestamp of
        the last line logged by \daemon{smtpd} for that connection.  If the
        \daemon{smtpd} timestamp is later there was a rejection between the
        accepted mail and the timeout, therefore the timeout applies to a
        rejected mail; the timeout has already been recorded so the timeout
        action finishes.  If the mail acceptance timestamp is greater then
        the timeout applies to the just-accepted mail, which will be
        discarded.

@@

This complication is further complicated by the presence of out of order \daemon{cleanup}
lines: see section~\refwithpage{discarding cleanup lines} for details.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances which trigger it.  Sometimes the
\daemon{cleanup} line is logged after the timeout line; parsing this line
causes the creation of a new state table entry for the queueid in the log
line.  This is incorrect because the line actually belongs to the mail
which has just been discarded, and the next log line for that queueid will
be seen when the queueid is reused, causing a queueid clash and the
appropriate warning.

In the case where the \daemon{cleanup} line is still pending the algorithm
updates a global cache of queueids, adding the queueid and the timestamp
from the timeout line.  When the next \daemon{cleanup} line is parsed for
that queueid the cache will be checked, and the line will be deemed part of
the discarded mail and discarded if it meets the following requirements:

@@

    \item The timestamp of the \daemon{cleanup} line must be within ten
        minutes of the mail acceptance timestamp.  Timeouts happen after
        five minutes, but some data may have been transferred slowly, and
        empirical evidence shows that ten minutes is not unreasonable;
        hopefully it is a good compromise between false positives and false
        negatives.

@@

The next \daemon{cleanup} line must meet the criteria above for it to be
discarded because not every connection where a timeout occurs will have a
\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
next \daemon{cleanup} line it would in some cases be mistaken.  Whether or
not the next \daemon{cleanup} line is discarded the queueid will be removed
from the cache of timeout queueids when the next \daemon{pickup} line
containing that queueid is parsed.

During the TIMEOUT action if the mail lacks a \daemon{cleanup} line, the
queueid and other data about the mail, particularly the timestamp of the
last log line, are added to a global cache of timeout queueids.  This cache
is subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action,
and if the queueid exists and the \daemon{cleanup} line passes the checks
listed above it will be discarded.

@@

Occasionally the \daemon{pickup} line logged when mail is submitted locally
via sendmail appears later in the logfile than the \daemon{cleanup} line
for that mail.  This seems to occur during periods of particularly heavy
load, so is most likely due to process scheduling vagaries.  Normally if
the queueid given in the \daemon{pickup} line exists a warning is generated
by the \daemon{pickup} action, but if the following conditions are met it
is assumed that the lines are out of order:

@@

    \item The only program which has logged anything thus far for the mail
        is \daemon{cleanup}.

    \item There is less than a five second difference in the timestamps of
        the \daemon{cleanup} and \daemon{pickup} lines.

@@

Occasionally a \daemon{smtpd} will just stop logging, without an
immediately obvious reason.  After poring over logs for some time there are
several reasons for this infrequent occurrence:

@@

    \item Postfix is stopped or its configuration is reloaded.  When this
        happens all \daemon{smtpd} processes exit, and all entries in the
        connections state table must be committed and deleted from the
        state table.  The queueid state table is untouched.

    \item Postfix sometimes forcibly kills a \daemon{smtpd}, so the active
        connection must be committed and deleted from the connections state
        table.

    \item Occasionally a \daemon{smtpd} will die or exit uncleanly, so the
        active connection must be committed and deleted from the
        connections state table.

@@

The above appear to cover all situations where a \daemon{smtpd} suddenly
stops logging.  In addition to removing an active connection the last
accepted mail may need to be discarded, as detailed in
section~\refwithpage{timeouts-during-data-phase}.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in section~\refwithpage{additional complications}
were detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
\pid{} is not found,\footnote{There will often be warnings about a missing
queueid or \pid{} in the first few hundred or thousand log lines because the
earlier log lines for those connections or mails are in the previous log
file; loading the saved state from the previous log file will solve this
problem.} or when there are problems tracking a child mail (see
section~\refwithpage{tracking re-injected mail}).  There should be few or
no warnings, and the state table should only contain entries for mails
which had yet to be delivered when the logfiles ended, or started before
the logfiles began.

@@

        Pflogsumm produces a report designed for troubleshooting, and does
        not support saving any data; parsing is fixed and concentrates on
        extracting only the data required to produce the report.  Both the
        parsing and reporting are difficult to extend, as it is a
        specialised tool.  It does not attempt to correlate log lines by
        queueid or \pid{}, nor does it need to deal with the complications
        encountered during this project.  Pflogsumm produces a useful
        report, and successfully dealt with all the Postfix logs tested
        with.

@@

        The interface is optimised for interactive rather than automated
        queries and it does not appear to be possible to write independent
        tools to utilise the Splunk database.  Splunk is a generic tool, so
        it lacks any Postfix specific support over and above extracting the
        \texttt{key=<value>} fields; most importantly it makes no attempt
        to correlate log lines by queueid or \pid{}, nor to handle any of
        the myriad complications discussed in this document.  Correlation
        of log lines is left up to the user, and Splunk certainly makes the
        job easier than using grep, awk and similar Unix tools unless the
        user has a very high proficiency with those tools.

@@

        Unlike other parsers reviewed, Lire does attempt to correlate log
        lines by queueid, but not by \pid{}, so e.g.\ the complete list of
        recipients for a mail should be available.
